---
title: "#SWDChallenge March 2019"
author: "Sam Robertson"
date: 2019-03-14
categories: []
tags: ["R", "data-visualization", "tidyverse"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

FILE_DATA <- "~/Data/swd-global-aid/aiddata-countries-only.csv"

data_raw <- 
  FILE_DATA %>% 
  read_csv(
    col_types = cols(
      X1 = col_integer(),
      aiddata_id = col_character(),
      aiddata_2_id = col_character(),
      year = col_integer(),
      donor = col_character(),
      recipient = col_character(),
      commitment_amount_usd_constant = col_double(),
      coalesced_purpose_code = col_integer(),
      coalesced_purpose_name = col_character()
    )
  )

data <-
  data_raw %>% 
  select(-X1, -aiddata_2_id) %>% 
  rename(
    id = aiddata_id,
    amount = commitment_amount_usd_constant,
    purpose_code = coalesced_purpose_code,
    purpose_desc = coalesced_purpose_name
  ) %>% 
  filter(amount >= 0)
```

This month I decided to take part in the [Storytelling With Data monthly challenge](http://www.storytellingwithdata.com/swdchallenge) for the first time! The dataset we were given to explore contains global aid exchanges between `r length(union(data %>% distinct(recipient) %>% pull(recipient), data %>% distinct(donor) %>% pull(donor)))` countries across the world across the years `r min(data %>% pull(year))`-`r max(data %>% pull(year))`. The goal is to create visualizations that answers the broad question: __Who Donates?__, as well as some bonus questions about distribution of donations geographically, temporally and by purpose of donation. Here's my initial attempt! Along with some code. (The only package I use here is `tidyverse`).

### The Data

The data provided is nice and clean, so all we are left to do is read it in using `read_csv()`. I changed some variable names to make it nicer to work with, and noticed that there are a few negative quantities of money in the data, which I drop since they are impossible. Here's a glimpse of what the data looks like:

```{r echo=FALSE}
data %>% head() %>% knitr::kable()
```

### Who Donates?

One of the challenges in answering this question is how to summarize across time. I chose to look at the proportion of the total money contributed to global aid that each country contributed and received.

```{r}
donated <-
  data %>% 
  group_by(donor) %>% 
  summarise(donated = sum(amount)) %>% 
  mutate(prop_donated = donated / sum(donated)) %>% 
  select(country = donor, prop_donated)

received <-
  data %>% 
  group_by(recipient) %>% 
  summarise(received = sum(amount)) %>% 
  mutate(prop_received = received / sum(received)) %>% 
  select(country = recipient, prop_received)

aid <-
  donated %>% 
  full_join(received, by = c("country")) %>% 
  mutate_at(vars(-country), ~if_else(is.na(.), 0, .)) %>% 
  gather(-country, key = direction, value = proportion_of_aid) %>% 
  mutate(direction = str_extract(direction, "[^_]+$"))

country_order <-
  aid %>% 
  spread(direction, proportion_of_aid) %>% 
  mutate(diff = donated - received) %>% 
  arrange(diff) %>% 
  pull(country)

aid <-
  aid %>% 
   mutate(country = factor(country, levels = country_order, ordered = TRUE))

segments <-
  aid %>% 
  spread(direction, proportion_of_aid)

aid %>% 
  ggplot(aes(y = country)) +
  geom_segment(
    aes(yend = country, x = donated, xend = received), 
    color = "grey40",
    data = segments
  ) +
  geom_point(aes(x = proportion_of_aid, color = direction), size = 2) +
  scale_y_discrete(expand = expand_scale(0)) +
  scale_x_sqrt(
    labels = scales::percent, 
    expand = expand_scale(0), 
    limits = c(0,0.4),
    breaks = c(0, 0.01, 0.025, 0.05, 0.1, 0.2, 0.3, 0.4)
  ) +
  scale_color_brewer(type = "qual", palette = "Set1", labels = str_to_title) +
  theme_minimal() +
  theme(legend.position = "top") +
  coord_cartesian(clip = "off") +
  labs(
    y = NULL,
    x = glue::glue("Percentage of Total Aid {min(pull(data, year))} - {max(pull(data, year))}"),
    color = NULL,
    title = "The United States and Japan are the world's major donors",
    subtitle = "India has received almost 40% of all global aid"
  )
```

Excuse the squished y-axis. I played around with it for a while and eventually gave up. Any hints are very welcome!

### Has the Amount Donated Changed Over Time?

In keeping with the same metric, proportion of aid contributed and received, we can also look at the trends over time. I've highlighted the top three donors and recipients in the figure. Interestingly, it seems that receiving tends to be steadier over time, while donations see more anomalous spikes

```{r}
donated <-
  data %>% 
  group_by(donor, year) %>% 
  summarise(donated = sum(amount)) %>% 
  ungroup() %>% 
  mutate(prop_donated = donated / sum(donated)) %>% 
  select(country = donor, year, prop_donated)

received <-
  data %>% 
  group_by(recipient, year) %>% 
  summarise(received = sum(amount)) %>% 
  ungroup() %>% 
  mutate(prop_received = received / sum(received)) %>% 
  select(country = recipient, year, prop_received)

timeseries <-
  donated %>% 
  full_join(received, by = c("country", "year")) %>% 
  mutate_at(vars(prop_donated, prop_received), ~if_else(is.na(.), 0, .)) %>% 
  gather(prop_donated, prop_received, key = direction, value = proportion) %>% 
  mutate(
    proportion = if_else(direction == "prop_donated", proportion, -proportion)
  )
  
top_3 <- 
  aid %>% 
  filter(direction == "donated") %>% 
  top_n(3, proportion_of_aid) %>% 
  pull(country)

bottom_3 <-
  aid %>% 
  filter(direction == "received") %>% 
  top_n(3, proportion_of_aid) %>% 
  pull(country)

timeseries_main <-
  timeseries %>%
  filter(country %in% top_3 & direction == "prop_donated" | 
           country %in% bottom_3 & direction == "prop_received")

country_order <- 
  timeseries_main %>% 
  filter(year == max(pull(data, year))) %>% 
  arrange(desc(proportion)) %>% 
  pull(country)

timeseries_main <-
  timeseries_main %>% 
  mutate(country = factor(country, levels = country_order, ordered = TRUE))

labeller <-
  function(y) {
    y = if_else(y < 0, -y, y)
    scales::percent(y)
  }

timeseries %>% 
  unite(group, country, direction, remove = FALSE) %>% 
  ggplot(aes(year, proportion)) +
  geom_line(aes(group = group), alpha = 0.2) +
  geom_line(aes(color = country), data = timeseries_main) +
  scale_x_continuous(breaks = seq(1970, 2015, by = 5)) +
  scale_y_continuous(labels = labeller) +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  theme_minimal() +
  labs(
    x = NULL,
    color = NULL,
    y = glue::glue("Percentage of Annual Aid"),
    title = "Contributions from significant donors is not constant over time",
    subtitle = "Major events like war and recession drive spikes in aid"
  ) +
  annotate(
    geom = "text",
    x = 1974,
    y = 0.0275,
    label = "Proportion Donated",
    hjust = 0
  ) +
  annotate(
    geom = "text",
    x = 1974,
    y = -0.0175,
    label = "Proportion Received",
    hjust = 0
  )
```

